\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-12-05 latex/paper.tex
feat: full review + conclusion draft

% author info
\author{Lionel J. Garcia}
\author{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}
\author{Dax L. Feliz}
\author{Catriona A. Murray}
\author{Francisco J. Pozuelos}
\correspondingauthor{Lionel J. Garcia}
\email{lgarcia@flatironinstitute.org}

@@

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with these nuisance signals, a common approach consists in cleaning a light curve from correlated noise before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signals down to no detection. As this degradation depends on the correlated noise characteristics, we explore the relative parameter space for which transits are altered, and quantify this effect on a set of simulated light curves. On this synthetic dataset, we show that \nuancemethod{} largely outperforms the detection capabilities of a commonly used transit search technique, which consists in cleaning light curves with an optimal bi-weight filter before searching for transits using a box-least-square algorithm. In order to assess the performance of \nuancemethod{} on real datasets, we inject and recover transits into the TESS light curves of 438 M dwarfs, and compare our method to 5 detrending techniques followed by a box-least-square search. For transits with a duration exceeding one fifth of the stellar variability timescale, i.e. when both signals start ressembling each other, we show that our algorithm is the most perforant in 94\% of cases, leading to both the highest number of true positives and the lowest number of false positives detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationaly expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices. Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of both active stars and their transiting exoplanets, both using space-based and sparse ground-based observations.
\end{abstract}

\section*{Introduction}
Transiting exoplanets are keystone objects for the field of exoplanetary science, but detecting transits in light curves featuring stellar variability and instrumental signals remains a challenge \citep{Pont2006,Howell2016}. For this reason, known transiting exoplanets tend to be found around quieter stars, or belong to the population of close-in giants whose transit signals dominate over stellar rotational variability \citep{Simpson2023}. However, discovering transiting exoplanets around active stars holds few promises. First, as younger stars are more active\footnote{approximately down to spectral type MV ()} \citep{Skumanich1972}, being able to detect planets transiting active stars will favor the discovery of young planetary systems \citep[e.g.][]{Newton2022}. Second, as stellar variability may originate from surface active regions (such as starpots), the increased discovery of occulting companions will increase our chances to map the photosphere of active stars \citep[e.g.][]{Morris2017}, benefiting both the study of stellar atmospheres and the concerning impact of their non-uniformity on planetary atmosphere retrievals \citep{rackham2018}. Overall, enabling the detection of transits in light curves with high levels of correlated noises will greatly benefit the study of terrestrial exoplanets around late M-dwarfs, usually observed at lower SNR and more likely to display photometric variability (e.g. \citealt{Murray2020}).
\\\\
Commonly used transit-search algorithms, such as the Box-Least-Square algorithm \citep[BLS,][]{bls} are capable of detecting transits in light curves containing only transit signals and white noise. Using this method, the simplest way to detect transits in a light curve featuring correlated noise (either astrophysical or instrumental), is to first clean it from nuisance signals before performing the search. This strategy is widely adopted by the community, both using physically-motivated systematic models like \cite{everest1, everest2}, or filtering techniques (\citealt{Jenkins2010}, \citealt{wotan}). However, when correlated noise starts resembling transits, this cleaning step (often referred to as \textit{detrending}) is believed to degrade their detectability \cite[see subsection 4.3 of][]{wotan}. In this case, the only alternative to search for transits is to perform a full-fledged modeling of the light curve, including both transits and correlated noise, and to compute the likelihood of the data to the transit model on a wide parameter space (an approach largely avoided due to its intractable nature). Nonetheless, \cite{kovacs2016} ask: \textit{Periodic transit and variability search with simultaneous systematics filtering: Is it worth it?}. The latter study explores a handful of cases and generally discards the benefit of using a full-fledged approach. However, it does not explore the light-curves characteristics for which such a complete modeling becomes necessary.\\\\
In this paper, we explore the light curve characteristics for which a full-fledge transit search is necessary, and we present \nuancemethod{}\footnote{Throughout the paper, \nuancemethod{} typeseted in italics refers to the method, while \nuancecode{} in sans-serif refers to its implementation.}, a method to search for transit signals while modeling correlated noises in a tractable way. In \autoref{issues}, we describe the effect of correlated noise on transit light curves and the effect of its detrending on planetary signals detectability. In \autoref{nuance}, we present \nuance{}. In \autoref{results}, we test the performance of \nuance{} on a wide variety of cases, and compare it to commonly used transit search algorithms. Finally, in \autoref{discussion}, we discuss the results and the limitations of \nuance{}, and we conclude in \autoref{conclusion}.

@@

A strong assumption when using BLS-based algorithms is that the searched dataset only contains transit signals and white noise. Hence, two sources of correlated noise particularly justify the need for a detrending step before searching for transits: instrumental signals (such as those induced by telescope pointing errors) and astrophysical signals (such as stellar variability induced by pulsations or starspots). In this section, we qualitatively study the effect of correlated noise and its detrending on transits detectability.

@@

One way to study the detectability of a unique transit signal is to compute its signal-to-noise-ratio (SNR, \citealt{Pont2006}, Equation 12):
\begin{equation}\label{eq:snr}
  SNR= \frac{\Delta}{\sqrt{\frac{\sigma_w^2}{n} + \frac{\sigma_c^2}{N_{tr}}}}
\end{equation}
where $\Delta$ is the relative transit depth, $n$ is the number of points within transit, $N_{tr}$ the number of transits ($N_{tr}=1$ here), and $\sigma_w^2$ and $\sigma_c^2$ are the white and correlated noise variances. \autoref{fig:issue1} shows this metric computed for a unique transit observed in the absence (grey) and presence (red) of correlated noise.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. A 1-hour transit signal of depth 1\% is generated on top of white noise (variance of $0.0015^2$) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a Gaussian process (GP) with a Matèrn-32 kernel of scale 1 hour and sigma of 0.2\%. The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{signals_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}

@@

@@

The presence of instrumental correlated noise motivated the development of systematics detrending algorithms, such as the Trend Filtering Algorithm (\textsc{TFA}, \citealt{tfa}, in its primary use case), \textsc{SysRem} (\citealt{sysrem}) or Pixel Level Decorrelation (\textsc{PLD}, \citealt{pld}; see also \textsc{Everest} from \citealt{everest1, everest2}). Most of these methods rely on the shared nature of instrumental signals among light-curves (or neighboring pixels) such that the correction applied should not degrade the transit signal and can be modeled using contemporaneous measurements (like detector's temperature, pointing error, sky background or airmass time series). In opposition, stellar variability and other astrophysical signals cannot be correlated with simultaneous measurements. This gave rise to several treatments in order to reconstruct and detrend stellar variability. One of them is physically-motivated and makes use of GP (e.g. \citealt{k2sc}), another is empirical and makes use of filtering algorithms (\citealt{Jenkins2010}, \citealt{wotan}).\\\\
In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different amplitudes and timescales, sampled from a GP with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same GP from which the data has been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that the degradation is strongly dependant on the correlated noise characteristics encountered.\\\\
In order to explore the parameter space for which detrending is the most problematic, we employ the method previously described to simulate 10 000 light curves containing a transit signal and correlated noise in the form of stellar variability. For each light curve, we reconstruct the variability signal and detrend it using an optimal bi-weight filter before estimating the remaining transit SNR. For convenience, the simulated light curves share a common transit added on top of white noise, and a variability signal only defined by two parameters: $\tau$, the relative timescale of the variability with respect to the transit duration; and $\delta$, the relative amplitude of the variability against transit depth, corresponding to an SHO kernel with hyperparameters chosen as: 

@@

For $\tau=1$ and $\delta=1$, the expressions of $\omega$ and $\sigma$ given in \autoref{eq:relative_params} correspond to a variability signal with a period half that of the transit duration, and a standard deviation two times that of the transit depth, i.e. strongly resembling the simulated transit signal.\\\\
\autoref{fig:snr_detrend} shows that it exists an entire region of the $(\tau, \delta)$ parameter space for which the bi-weight detrending degrades transit SNR to the point of no detection ($SNR < 6$). Hence, bi-weight filter detrending makes transit-search blind to many systems. While this study should be extended to other detrending techniques, it highlights the need for a more informed transit search algorithm able to deal with correlated noise.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../workflows/cleaning_snr/figures/result.pdf}
        \caption{SNR of a unique transit after detrending light curves using an optimal bi-weight filter. All light curves correspond to a 2.8 days observation with a cadence of 2 minutes, and contain a unique transit of duration 1 hour and a depth of 1\%, added on top of white noise with a standard deviation of 0.5\% plus stellar variability. Light curves at the top and right side of the central plot are shown with their corresponding $\tau$ and $\delta$ values. The amplitude of the variability increases with $\delta$ and the timescale of the variability increases with $\tau$.}
        \label{fig:snr_detrend}
    \end{centering}
\end{figure}

@@

\textit{nuance} is an algorithm capable of searching for planetary transits in light curves containing correlated noise, such as instrumental signals and photometric stellar variability.
\\\\
Let assume that the flux $f$ of a star is observed and arranged in the vector $\bm{f}$ of size $N$, associated to the vector of times $\bm{t}$. This flux, shown in \autoref{fig:linear_search}, contains instrumental signals, stellar variability and a periodic transit signal that we wish to uncover. We assume that a set of $M$ observed measurements (such as the position of the star on the detector or the sky background) taken at the same time as the flux can be treated as explanatory variables for $\bm{f}$. These measurements constitute the columns of the $(N\times M)$ \textit{design matrix} $\bm{X}$.\\\\

Ideally, we would detect the periodic transit signal in this flux by sampling the posterior likelihood of this data to a full-fledge model including stellar variability (more generally correlated noise), instrumental systematic signals (modeled with explanatory variables or treated as correlate noise), and a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $\Delta$. We would then reduce the posterior likelihood to $p(\bm{f}\vert P)$, its marginalized version over all parameters except the period $P$, producing a transit search periodogram $\mathcal{Q}(P)$. However, this approach has two issues: It is highly intractable, and it may lead to multimodal distributions that are hard to interpret.
\\\\
Given a period $P$, we instead want to compute the likelihood of a periodic transit signal at the maximum likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, i.e the periodogram
\begin{equation}\label{eq:periodogram}
        \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)
\end{equation}
This is done by adopting the strategy of \cite{foreman2016}, and separate the transit search into two components: the \textit{linear search} and the periodic search. During the \textit{linear search}, the likelihood of a single non-periodic transit is computed for a grid of epochs, durations and depths. Then, the periodic search consists in combining these likelihoods to compute the likelihood of the data given a periodic transit signal for a range of periods. These combined likelihoods yield a transit-search periodogram on which the periodic transit detection is based. \textsf{nuance} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a GP, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \textsf{nuance} searches for transits while, at the same time, modeling correlated noise, avoiding the detrending step that degrades transit signals SNR.\\\\
We note that the approach employed by \textsf{nuance} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

During the \textit{linear search}, the goal is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of the data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noise, the light curve $f$ is modeled as being drawn from a GP, like the simulated light curves used in \autoref{issues}, such that
\begin{equation*}
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
\end{equation*}
with mean $\bm{Xw}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, this signal is computed and appended as the last column of the design matrix $\bm{X}$, using the simple transit model from \cite{protopapas} with a unitary depth (\autoref{eq:protopapas}). This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi,
\end{equation}
where the parameters vector $\bm{w}$ and their errors $\bm{\sigma}$ are computed using the generalized least-square solution
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{and} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1},
\end{equation} 
with $\bm{\Sigma}$ the covariance matrix modeled using the GP. Hence, $\ln p(\bm{f} \vert I)$ can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. In this equation. This \textit{linear search} leads to the set of likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})}_{i, j},
\end{equation*}
were $\Delta_{i,j}$ is the depth linearly solved for a given $(T_i, D_j)$\footnote{The expression of $\set{\ln\mathcal{L}_{i,j}}_{i, j}$ omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{i,j}$) as it is also linearly solved for any given value of $(T_i, D_j)$ and irrelevant in what follows.}. \autoref{fig:linear_search} shows this likelihood grid computed for a simulated dataset, using the same GP  and design matrix $\bm{X}$ used to simulate the data.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_linear_search.pdf}
        \caption{Principle and output of the \textit{linear search}. The simulated dataset (top) corresponds to the one shown and described in \autoref{fig:app_principle_dataset}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of indices $(i,j)$, the likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors inferred linearly using \autoref{eq:ls}.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

nu = Nuance(time, flux, gp=gp, X=X)
\end{lstlisting}
where \textsf{gp} is a \textsf{tinygp} GP instance and \textsf{X} the design matrix of the linear model. \textsf{nuance} exploits the use of \textsf{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}}, a Python package powered by \textsf{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}, allowing for custom kernels to be built and highly tractable computations. We can then define a set of epochs \textsf{t0s} and durations \textsf{Ds} and run the \textit{linear search} with
\begin{lstlisting}[language=Python,linewidth=\linewidth]
import numpy as np

>>>>

% author info
\author{Lionel J. Garcia}
\author{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}
\author{Dax L. Feliz}
\author{Catriona A. Murray}
\correspondingauthor{Lionel J. Garcia}
\email{lgarcia@flatironinstitute.org}

@@

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with them, a common approach consists in cleaning a light curve from nuisance signals before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signal-to-noise-ratio down to no detection. As this degradation depends on the correlated noise characteristics, we explore the relative parameter space for which transits are altered, and quantify this effect on a set of simulated light curves. On this synthetic dataset, we show that \nuancemethod{} largely outperforms the detection capabilities of a commonly used transit search technique, which consists in cleaning light curves from styellar variability with an optimal bi-weight filter before searching for transits using a box-least-square algorithm. Then, in order to assess the performance of \nuancemethod{} on real datasets, we inject and recover transits into the TESS light curves of 438 M dwarfs, and compare our method to 4 detrending techniques followed by a Box-Least-Square search. For transits with a duration exceeding one fifth of the stellar variability timescale, i.e. when both signals start resembling each other, we show that our algorithm is the most performant in 94\% of cases, leading to both the highest number of true positives and the lowest number of false positive detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationaly expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices (?). Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of active stars and their transiting exoplanets, both using space-based and sparse ground-based observations.
\end{abstract}

\section*{Introduction}
Transiting exoplanets are keystone objects for the field of exoplanetary science, but detecting transits in light curves featuring stellar variability and instrumental signals remains a challenge \citep{Pont2006,Howell2016}. For this reason, known transiting exoplanets tend to be found around quieter stars, or belong to the population of close-in giants whose transit signals dominate over stellar rotational variability \citep{Simpson2023}. However, discovering transiting exoplanets around active stars holds few promises. First, as younger stars are more active \citep{Skumanich1972}, being able to detect planets transiting active stars will favor the discovery of young planetary systems \citep[e.g.][]{Newton2022}. Second, as stellar variability may originate from surface active regions (such as starpots), the increased discovery of occulting companions will increase our chances to map the photosphere of active stars \citep[e.g.][]{Morris2017}, benefiting both the study of stellar atmospheres and the concerning impact of their non-uniformity on planetary atmosphere retrievals \citep{rackham2018}. Overall, enabling the detection of transits in light curves with high levels of correlated noises will greatly benefit the study of terrestrial exoplanets around late M-dwarfs, usually observed at lower SNR and more likely to display photometric variability (e.g. \citealt{Murray2020}).
\\\\
Commonly used transit-search algorithms, such as the Box-Least-Square algorithm \citep[BLS,][]{bls} are capable of detecting transits in light curves containing only transit signals and white noise. Using this method, the simplest way to detect transits in a light curve featuring correlated noise (either astrophysical or instrumental), is to first clean it from nuisance signals before performing the search. This strategy is widely adopted by the community, both using physically-motivated systematic models like \cite{everest1, everest2}, or filtering techniques (\citealt{Jenkins2010}, \citealt{wotan}). However, when correlated noise starts resembling transits, this cleaning step (often referred to as \textit{detrending}) is believed to degrade their detectability \cite[see subsection 4.3 of][]{wotan}. In this case, the only alternative to search for transits is to perform a full-fledged modeling of the light curve, including both transits and correlated noise, and to compute the likelihood of the data to the transit model on a wide parameter space (an approach largely avoided due to its intractable nature). Nonetheless, \cite{kovacs2016} ask: \textit{Periodic transit and variability search with simultaneous systematics filtering: Is it worth it?}. The latter study explores a handful of cases and generally discards the benefit of using a full-fledged approach. However, it does not explore the light-curves characteristics for which such a complete modeling becomes necessary.\\\\
In this paper, we explore the light curve characteristics for which a full-fledge transit search is necessary, and we present \nuancemethod{}\footnote{Throughout the paper, \nuancemethod{} written in italics refers to the algorithmic and analytical method, while \nuancecode{} in sans-serif refers to its implementation.}, a method to search for transit signals while modeling correlated noises in a tractable way. In \autoref{issues}, we describe the effect of correlated noise on transit light curves and the effect of its detrending on transit signals detectability. In \autoref{nuance}, we present \nuance{}, and the two main steps on which this method is based: the linear search and the periodic search. In \autoref{results}, we test the performance of \nuance{} on a wide variety of cases, and compare it to commonly used transit search algorithms. This include transits injected in synthetic datasets, but also in the TESS light curves of 438 rapidly-rotating M dwarfs. Finally, in \autoref{discussion}, we discuss the results and the limitations of \nuance{}, and we conclude in .

@@

A strong assumption when using the BLS algorithm to search for transits is that the searched dataset only contains these signals in addition of white noise. Hence, two sources of correlated noise particularly justify the need for a detrending step before searching for transits: instrumental signals (such as those induced by telescope pointing errors) and astrophysical signals (such as stellar variability induced by pulsations or starspots). In this section, we qualitatively study the effect of correlated noise and its detrending on transits detectability.

@@

One way to study the detectability of a unique transit signal is to compute its signal-to-noise-ratio (SNR, \citealt{Pont2006}, Equation 12):
\begin{equation}\label{eq:snr}
  SNR= \frac{\Delta}{\sqrt{\frac{\sigma_w^2}{n} + \frac{\sigma_c^2}{N_{tr}}}}
\end{equation}
where $\Delta$ is the relative transit depth, $n$ is the number of points within transit, $N_{tr}$ the number of transits ($N_{tr}=1$ here), and $\sigma_w^2$ and $\sigma_c^2$ are the white and correlated noise variances\footnote{i.e. $\sigma_w^2 = \sum_{i=j} C_{{ij}}$ and $\sigma_c^2 = \sum_{i\neq j} C_{{ij}}$ with $C$ the covariance matrix of the data.}. \autoref{fig:issue1} shows this metric computed for a unique transit observed in the absence (grey) and presence (red) of correlated noise.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. A 1-hour transit signal of depth 1\% is generated on top of white noise (variance of $0.0015^2$) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a Gaussian process (GP; \citealt{Rasmussen2005}) with a Matèrn-32 kernel of scale 1 hour and sigma of 0.2\%. The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{signals_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}

@@

The presence of instrumental correlated noise motivated the development of systematics detrending algorithms, such as the Trend Filtering Algorithm (\textsc{TFA}, \citealt{tfa}, in its primary use case), \textsc{SysRem} (\citealt{sysrem}) or Pixel Level Decorrelation (\textsc{PLD}, \citealt{pld}; see also \textsc{Everest} from \citealt{everest1, everest2}). Most of these methods rely on the shared nature of instrumental signals among light-curves (or neighboring pixels) such that the correction applied should not degrade the transit signal and can be modeled using contemporaneous measurements (like detector's temperature, pointing error, sky background or airmass time series).

@@

In opposition, stellar variability and other astrophysical signals cannot be correlated with simultaneous measurements. This gave rise to several treatments in order to reconstruct and detrend stellar variability. Some of them are physically-motivated and make use of GPs (e.g. \citealt{k2sc}), others are empirical and make use of filtering and data-driven algorithms (\citealt{Jenkins2010}, \citealt{wotan}).\\\\
In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different amplitudes and timescales, sampled from a GP with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same GP from which the data has been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that this degradation due to detrending is strongly dependant on the correlated noise characteristics encountered.\\\\
In order to explore the parameter space for which detrending is the most problematic, we employ the method previously described to simulate 10 000 light curves containing a transit signal and correlated noise in the form of stellar variability. For each light curve, we reconstruct the variability signal and detrend it using an optimal bi-weight filter, i.e. with a window size three times that of the transit duration \citep{wotan}, before estimating the remaining transit SNR. For convenience, the simulated light curves share a common transit added on top of white noise, and a variability signal only defined by two parameters: $\tau$, the relative timescale of the variability with respect to the transit duration; and $\delta$, the relative amplitude of the variability against transit depth, corresponding to an SHO kernel with hyperparameters chosen as: 

@@

For $\tau=1$ and $\delta=1$, the expressions of $\omega$ and $\sigma$ given in \autoref{eq:relative_params} correspond to a variability signal with a period half that of the transit duration, and a correlated noise amplitude comparable to the transit depth, i.e. strongly resembling the simulated transit signal.\\\\
\autoref{fig:snr_detrend} shows that it exists an entire region of the $(\tau, \delta)$ parameter space for which the bi-weight detrending degrades transit SNR to the point of no detection ($SNR < 6$). Hence, bi-weight filter detrending makes transit search blind to many systems. While this study should be extended to other detrending techniques, it highlights the need for a more informed transit search algorithm able to deal with correlated noise, at least if present in the form of stellar variability.

% ## STOPPED here

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../workflows/cleaning_snr/figures/result.pdf}
        \caption{SNR of a unique transit after detrending light curves using an optimal bi-weight filter (i.e.\;of window size $3\times D$). All light span 2.8 days with exposure times of 2 minutes. The 10 000 simulated datasets contain the same transit signal with a duration $D=1$ hour and a depth of 1\%, added on top of white noise with a standard deviation of 0.5\% plus stellar variability. Light curves at the top and right side of the central plot are shown with their corresponding $\tau$ and $\delta$ values, which again correspond to the relative timescale of stellar variability against the transit duration, and the relative amplitude of stellar variability against the transit depth. Hence, the amplitude of the variability increases with $\delta$ and the timescale of the variability increases with $\tau$, the transit signal remaining fixed.}
        \label{fig:snr_detrend}
    \end{centering}
\end{figure}

@@

\textit{nuance} is an algorithm capable of searching for planetary transits in light curves containing correlated noise, such as instrumental signals and photometric stellar variability.
\\\\
Let assume that the flux $f$ of a star is observed and arranged in the vector $\bm{f}$ of size $N$, associated to the vector of times $\bm{t}$. This flux, shown in \autoref{fig:linear_search}, contains instrumental signals, stellar variability and a periodic transit signal that we wish to detect. We assume that a set of $M$ observed measurements (such as the position of the star on the detector or the sky background) taken at the same time as the flux can be treated as explanatory variables for $\bm{f}$. These measurements constitute the columns of the $(N\times M)$ \textit{design matrix} $\bm{X}$.\\\\
Ideally, we would detect the periodic transit signal in this flux by sampling the posterior likelihood of this data to a full-fledge model including stellar variability (more generally correlated noise), instrumental systematic signals (modeled with explanatory variables or treated as correlate noise), and a periodic transit signal of period $P$, epoch $T_0$, duration $D$ and depth $\Delta$. We would then reduce the posterior likelihood to $p(\bm{f}\vert P)$, its marginalized version over all parameters except the period $P$, producing a transit search periodogram $\mathcal{Q}(P)$. However, this approach has two issues: It is highly intractable, and it may lead to a multimodal distribution that is hard to interpret.
\\\\
Given a period $P$, we instead want to compute the likelihood of a periodic transit signal at the maximum likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, i.e the periodogram
\begin{equation}\label{eq:periodogram}
        \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)
\end{equation}
This is done by adopting the strategy of \cite{foreman2016}, separating the transit search into two components: the linear search and the periodic search. During the linear search, the likelihood of a single non-periodic transit is computed for a grid of epochs, durations and depths. Then, the periodic search consists in combining these likelihoods to compute the likelihood of the data given a periodic transit signal for a range of periods. These combined likelihoods yield a transit-search periodogram on which the periodic transit detection is based. \textsf{nuance} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a GP, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \textsf{nuance} searches for transits while, at the same time, modeling correlated noise, avoiding the detrending step that degrades transit signals SNR.\\\\
We note that the approach employed by \nuancemethod{} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

The goal of the linear search is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of the data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noise, the light curve $f$ is modeled as being drawn from a GP, like the simulated light curves used in \autoref{issues}, such that
\begin{equation*}
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
\end{equation*}
with mean $\bm{Xw}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, this signal is computed and appended as the last column of the design matrix $\bm{X}$, using the simple transit model from \cite{protopapas} with a unitary depth (\autoref{eq:protopapas}). This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi,
\end{equation}
where the parameters vector $\bm{w}$ and their errors $\bm{\sigma}$ are computed using the generalized least-square solution
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{and} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1},
\end{equation} 
with $\bm{\Sigma}$ the covariance matrix modeled using the GP. Hence, $\ln p(\bm{f} \vert I)$ can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. This \textit{linear search} leads to the set of likelihoods
\begin{equation*}
    \set{\ln\mathcal{L}_{i,j}}_{i, j} = \set{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})}_{i, j},
\end{equation*}
were $\Delta_{i,j}$ is the depth linearly solved for a given $(T_i, D_j)$\footnote{The expression of $\set{\ln\mathcal{L}_{i,j}}_{i, j}$ omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{i,j}$) as it is also linearly solved for any given value of $(T_i, D_j)$ and irrelevant in what follows.}. \autoref{fig:linear_search} shows this likelihood grid computed for a simulated dataset, using the same GP  and design matrix $\bm{X}$ used to simulate the data.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{../workflows/principle/figures/principle_linear_search.pdf}
        \caption{Principle and output of the linear search. The simulated dataset (top) corresponds to the one shown and described in \autoref{fig:app_principle_dataset}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of indices $(i,j)$, the likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors inferred linearly using \autoref{eq:ls}.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

@@

nu = Nuance(time, flux, gp=gp, X=X)
\end{lstlisting}
where \textsf{gp} is a \textsf{tinygp} GP instance and \textsf{X} the design matrix of the linear model. \textsf{nuance} exploits the use of \textsf{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}}, a Python package powered by \textsf{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}, allowing for custom kernels to be built and highly tractable computations. We can then define a set of epochs \textsf{t0s} and durations \textsf{Ds} and run the linear search with
\begin{lstlisting}[language=Python,linewidth=\linewidth]
import numpy as np
